---
output:
  html_document:
    keep_md: yes
---
=============================================================================

# Weather Events Harmful to the General Population and Economy

## Edward Lee 7/27/2014

### Course Class Reproducible Research Assignment 27/2014

From the NOAA weather data, we seek to answer which weather event will affect the general population
in terms of human lives as well as the economic damages.  The data will be from the from NOAA weather 
[storm database](http://www.ncdc.noaa.gov/stormevents/details.jsp).
**.pdf** documentation is at [NOAA](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf).

Simply put, across the United States, which types of events are most harmful with respect 
to population health?  Which types of events have the greatest economic consequences?

## Data Processing

There will be categorization of event type that may be subjective.   Not to mention the data collection
of the weather events will be categorized on the field manually and hence subjected to human input errors.

```{r Set my working Dir}
setwd ("C:/Cousera/Course5_Reproducible Research")
```

```{r Downloading Data from Internet}
setInternet2(TRUE)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,destfile,mode="wb")

url2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
destfile2 <- "repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
download.file(url2,destfile2,mode="wb")
```

```{r Read the file in BZ format - cool! - and then check basic EDA}
StormData<- read.csv(bzfile(destfile), stringsAsFactors = FALSE)
names(StormData)
summary(StormData$FATALITIES)
summary(StormData$INJURIES)
```

It seems that there are 985 unique types of events, I will endeavor to decrease this list a lot more
I will also create a list for Question 1 for most harmful and Question 2 for damages PROPDMGEXP,CROPDMGEXP
Also, I will only need certain columns FATALITIES INJURIES PROPDMG CROPDMG PROPDMGEXP CROPDMGEXP

``` {r Keeping on the needed columns and Row using SUBSET}
KeepCol <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG","CROPDMG","PROPDMGEXP","CROPDMGEXP")
StormDataKeep <- subset(StormData, select = KeepCol)
names(StormDataKeep)
StormDataKeepNZ <- subset(StormDataKeep, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)
```


```{r New Rollup the EVTYPE - Fix the actual numbers as well}
StormDataKeepNZ$EVENT <- StormDataKeepNZ$EVTYPE
StormDataKeepNZ$EVENT[grep("flood",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("warm",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HEAT"
StormDataKeepNZ$EVENT[grep("freeze",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("avala",StormDataKeepNZ$EVENT, ignore.case = T)] <-"AVALANCHE"
StormDataKeepNZ$EVENT[grep("ice",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("bitter",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("BELOW NORMAL PRECIPITATION",StormDataKeepNZ$EVENT, ignore.case = T)] <-"DROUGHT"
StormDataKeepNZ$EVENT[grep("BLIZZARD",StormDataKeepNZ$EVENT, ignore.case = T)] <-"SNOW"
StormDataKeepNZ$EVENT[grep("SNOW",StormDataKeepNZ$EVENT, ignore.case = T)] <-"SNOW"
StormDataKeepNZ$EVENT[grep("dry",StormDataKeepNZ$EVENT, ignore.case = T)] <-"DROUGHT"
StormDataKeepNZ$EVENT[grep("BEACH",StormDataKeepNZ$EVENT, ignore.case = T)] <-"BEACH EROSION"
StormDataKeepNZ$EVENT[grep("FIRE",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FIRE"
StormDataKeepNZ$EVENT[grep("COLD",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("RAIN",StormDataKeepNZ$EVENT, ignore.case = T)] <-"RAIN"
StormDataKeepNZ$EVENT[grep("CHILL",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("wet",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("frost",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("freez",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("gust",StormDataKeepNZ$EVENT, ignore.case = T)] <-"WIND"
StormDataKeepNZ$EVENT[grep("hail",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HAIL"
StormDataKeepNZ$EVENT[grep("heat",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HEAT"
StormDataKeepNZ$EVENT[grep("hurricane",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HURRICANE"
StormDataKeepNZ$EVENT[grep("typhoon",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HURRICANE"
StormDataKeepNZ$EVENT[grep("ice",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("icy",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("LANDSLIDE",StormDataKeepNZ$EVENT, ignore.case = T)] <-"LANDSLIDES"
StormDataKeepNZ$EVENT[grep("mud",StormDataKeepNZ$EVENT, ignore.case = T)] <-"MUDSLIDES"
StormDataKeepNZ$EVENT[grep("high temp",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HEAT"
StormDataKeepNZ$EVENT[grep("TROPICAL STORM",StormDataKeepNZ$EVENT, ignore.case = T)] <-"TROPICAL STORM"
StormDataKeepNZ$EVENT[grep("light",StormDataKeepNZ$EVENT, ignore.case = T)] <-"THUNDERSTORM"
StormDataKeepNZ$EVENT[grep("tstm",StormDataKeepNZ$EVENT, ignore.case = T)] <-"THUNDERSTORM"
StormDataKeepNZ$EVENT[grep("torn",StormDataKeepNZ$EVENT, ignore.case = T)] <-"TORNADO"
StormDataKeepNZ$EVENT[grep("tide",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("tsu",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("thun",StormDataKeepNZ$EVENT, ignore.case = T)] <-"THUNDERSTORM"
StormDataKeepNZ$EVENT[grep("tsu",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"


StormDataKeepNZ$PropertyDamageAmt <- StormDataKeepNZ$PROPDMG
MultLookup <- c(M = 10^6, m = 10^6, K = 10^3, k = 10^3, B = 10^9, b = 10^9)

StormDataKeepNZ$PropertyDamageAmt <- StormDataKeepNZ$PROPDMG * MultLookup [as.character(StormDataKeepNZ$PROPDMGEXP)]
StormDataKeepNZ$CropDamageAmt     <- StormDataKeepNZ$CROPDMG * MultLookup [as.character(StormDataKeepNZ$CROPDMGEXP)]
StormDataKeepNZ$TotDamageAmt   <- StormDataKeepNZ$PropertyDamageAmt + StormDataKeepNZ$CropDamageAmt
```

====================================================================

# RESULTS
Now we will aggregate the results and output the top 10 events graphically.  This will answer the
2 questions

``` {r Aggregate and Graph - somehow the aggregate function does not work - using plyr - get top 10}
library(plyr)
SFatal <- ddply(StormDataKeepNZ, .(EVENT), summarize, DEATHS = sum(FATALITIES, na.rm = TRUE))
SInjury <- ddply(StormDataKeepNZ, .(EVENT), summarize, INJURED = sum(INJURIES, na.rm = TRUE))
SPropDam <- ddply(StormDataKeepNZ, .(EVENT), summarize, PROPERTYCOST = sum(PropertyDamageAmt, na.rm = TRUE))
SCropDam <- ddply(StormDataKeepNZ, .(EVENT), summarize, CROPCOST = sum(CropDamageAmt, na.rm = TRUE))
STotDam <- ddply(StormDataKeepNZ, .(EVENT), summarize, TOTALCOST = sum(TotDamageAmt, na.rm = TRUE))


SDEATH10 <- head(SFatal[order(-SFatal$DEATHS),],10)
SINJURED10 <- head(SInjury [order(-SInjury$INJURED),],10)
SPropDam10 <- head(SPropDam[order(-SPropDam$PROPERTYCOST),],10)
SCropDam10 <- head(SCropDam[order(-SCropDam$CROPCOST),],10)
STotDam10 <- head(STotDam[order(-STotDam$TOTALCOST),],10)

SDEATH10
SINJURED10
SPropDam10
SCropDam10
STotDam10

```
=========================================================

## Graphing the results

```{r results1,fig.height=20,fig.width=20,cache=TRUE}
library (ggplot2)
ggplot (SDEATH10,aes(EVENT)) + 
	geom_histogram(aes(weight=DEATHS, fill =..count..),binwidth=1) + 
	xlab("EVENTS") + theme(axis.text.x=element_text(angle = 90)) + 
	ylab("FATALITIES") + 
	ggtitle("FATALITIES BY EVENTS")
```

```{r results2,fig.height=20,fig.width=20,cache=TRUE}
ggplot (SINJURED10,aes(EVENT)) + 
	geom_histogram(aes(weight=INJURED, fill =..count..),binwidth=1) + 
	xlab("EVENTS") + theme(axis.text.x=element_text(angle = 90)) + 
	ylab("INJURIES") + 
	ggtitle("INJURIES BY EVENTS")
```

```{r results3,fig.height=20,fig.width=20,cache=TRUE}
ggplot (STotDam10,aes(EVENT)) + 
	geom_histogram(aes(weight=TOTALCOST, fill =..count..),binwidth=1) + 
	xlab("EVENTS") + theme(axis.text.x=element_text(angle = 90)) + 
	ylab("DAMAGE COST") + 
	ggtitle("TOTAL DAMAGE COST")
```

	
================================================================================

## SUMMARY

The main causes of property damage are floods, hurricanes and toradoes
Crop damages are generally caused by drought, flood, and cold

Fatalities are far and away caused by tornadoes and thunderstorms
Injuries are also far and away caused by tornadoes, with heat, thunderstorms.



